Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2014-10-06 article.Rmd
Working on conclusion

One 2013 analysis of 2,500 recent history Ph.D.'s found that "gender played little role in employment patterns across particular professions and industries" [Wood and Townsend 2013]. Our own analysis reveals that the story is different for book publishing than for employment. The flagship journal of professional historians continues to publish twice as many reviews of books by male authors as female authors. This disparity is put into even sharper relief when set against the relative parity achieved by women in producing history dissertations. Women might be getting hired at comparable rates to men, but discriminatory gatekeeping remains alive and well in the profession.

Our program allows us to quantitatively measure the scope of the problem on a large scale and how it has changed over time. [to write later: include thoughts how this is even more important as gender discrimination turns away from the out-in-the-open kind of discrimination to more subtle disadvantages and systemic problems.]

@@

Our algorithm for predicting gender from first names, when combined with historical datasets in the Gender package, provides a powerful new method for researchers interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance many humanists continue to deal with the problem of scarcity in their sources. Our method uses the abundant existing datasets to infer more information from scarce datasets. The temporal dimension of our work matches historical questions to a historical method, providing much improved results over simple, anachronistic lookup methods. This method is extensible to any time or place for which there is a suitable dataset of names from which to make predictions about gender. This method is potentially useful to anyone who has a dataset of names, especially if those names can be associated with particular dates. As we have demonstrated in our study of gatekeeping in the American historical profession, our method can reveal important insights about gender even from datasets which do not otherwise include that information, and it can do so over long historical periods.

@@

@@

>>>>

One 2013 analysis of 2,500 recent history Ph.D.'s found that "gender played little role in employment patterns across particular professions and industries" [Wood and Townsend 2013]. Our own analysis reveals that the story is different for book publishing than for employment. The flagship journal of professional historians continues to publish twice as many reviews of books by male authors as female authors. This disparity is put into even sharper relief when set against the relative parity achieved by women in producing history dissertations. Women might be getting hired at comparable rates to men, but discriminatory gatekeeping remains alive and well in the profession. As explicit instances of sexism fade from view the disadvantages women face are growing more and more opaque.  But that does not make them any less damaging. As discrimination becomes less blatant and more systemic, digital methods like the Gender package are a way to a shine light on these cloaked disadvantages, to reveal the true extent of the problem and the work that remains to be done.

@@

Our algorithm for predicting gender from first names, when combined with historical datasets in the Gender package, provides a powerful new method for researchers interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. The temporal dimension of our work matches historical questions to a historical method, providing much improved results over simple, anachronistic, or ahistorical lookup methods. This method is also extensible to different times and places, provided a researcher has a similar reference dataset of names and genders from the period or area under study.

The analytical category of gender has transformed research in the humanities for the past half-century. More recently, the digitization movement has opened up exciting opportunities to interrogate gender in new ways. Projects in the digital humanities have ranged from recovering the voices of early-modern women writers to tracing the linguistic styles of male and female authors [Women Writers Project n.d., Argamon et. al. 2009]. The Gender package is a new tool to add to this approach. It reverses the traditional disciplinary direction bewteen computer science and the humanities. Rather than adapting existing computational methodologies, the Gender package uses a humanistic starting point---the historical relationship between gender and naming practices---to contribute a new method to computer science.

@@

[Argamon et. al. 2009] Argamon, Shlomo, Jean-Baptiste Goulain, Russell Horton, and Mark Olsen. "Vive la Différence! Text Mining Gender Difference in French Literature" *Digital Humanities Quarterly*. 3.2 (2009).

@@

[Women Writers Project n.d.] Women Writers Project. "About." <http://www.wwp.northeastern.edu/about/>. Accessed 5 October 2014.
